---
title: "Packages and functions"
output: html_notebook
---

# Packages

*All* the packages needed for this project should be loaded here so they can be
easily reloaded after a restart.

```{r}
# loading tidyverse loads ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr 
# and forcats
library("tidyverse")
library("nlme") # gls regression
```

# Functions

## Harmonise police force name

```{r}
harmonise_force <- function (force_name) {
  force_name %>% 
    str_replace_all(" & ", " and ") %>% 
    str_to_title() %>% 
    str_replace_all("And", "and") %>% 
    str_replace_all("Of", "of") %>% 
    str_replace("London, City of", "City of London") %>% 
    str_replace("Police", "")
}
```

## Test minimum findable effect size for a variable in an OLS model

  1. Specify a base model without the term of interest for comparison purposes.
  2. Construct a dataset with a known relationship between the response variable
     and the variable of interest, incorporating the known variance of that
     variable, e.g. specify that an extra 100 officers will lead to a certain
     decrease in the accident rate and then construct the variable based on 
     that.
  3. Run that model and use the formula at https://www.statmethods.net/stats/power.html
     to calculate f^2 using the differences between R^2 for the two models.
  4. Use f^2 to calculate the minimum sample size needed to detect an effect of
     that size using `pwr::pwr.f2.test()`. Remember that `v` in that function is
     the denominator degrees of freedom and sample size `n = v + u + 1` as 
     explained by [Ford (2018)](https://cran.r-project.org/web/packages/pwr/vignettes/pwr-vignette.html).

All this is likely to be easier if the continuous predictors are both scaled and 
mean-centered using `scale()`.

An alternative method to consider is outlined from page 10 
of [Blomberg 2014](https://pdfs.semanticscholar.org/2b85/0bb035e93663f835a5453c4f02ae55ac65da.pdf)
which would use repeated `anova()` tests to simulate the likelihood of a given
effect size producing a model that was a significant improvement on the base
model produced at step 1 (above).

The result of this would be something like "a one-SD increase in police officers
(X officers in the average force) would have to be associated with a Y change 
[or possibly 'decrease' — think if one- or two-tailed is needed] in accidents in
order to have an 80% chance of being detected using this model."
