---
title: "Data cleaning and formating"
output: html_notebook
---

# Data sources

The data used in this study are from:

  * **collisions data** from [Road Accident and Safety Statistics Table RAS50016](https://www.gov.uk/government/statistical-data-sets/ras50-contributory-factors)
  * **police workforce data** and **police functions data** from [HMICFRS Value for Money data](https://www.justiceinspectorates.gov.uk/hmicfrs/data/value-for-money-data/)
  * **historical police workforce data**
  * **vehicle kilometres data** from [Road Traffic Statistics Table TRA8904](https://www.gov.uk/government/statistical-data-sets/tra89-traffic-by-local-authority) 

All data are available under [version 3 of the Open Government Licence](https://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/). 
This licence explicitly allows relicensing of further-processed data under a 
CC-BY licence.

## Subsidiary data

Other data needed for this project:

  * [Lookup table between police-force areas and local authorities](http://geoportal.statistics.gov.uk/datasets/local-authority-district-to-community-safety-partnerships-to-police-force-areas-december-2016-lookup-in-england-and-wales)


# Pre-processing

Since the government typically does not provide data tables in machine-readable 
formats, the files have been lightly pre-processed before importing into R.
This includes removing footnotes and multiple header rows, as well as separating
multi-worksheet spreadsheets into flat CSV files. All other processing (e.g.
wide-to-long conversion) is done in this file.


# Police workforce and functions data

These data come from HMICFRS, which produces data on both police workforce and 
officer functions. The functions data are consistent over time, whereas the 
corresponding Home Office data used different definitions from 2015 onwards
(see the [User Guide to Police Workforce Statistics](https://www.gov.uk/government/publications/user-guide-to-police-workforce-statistics)). Officers are considered to be allocated to traffic if they appear in the
'Roads Policing' category of the HMICFRS data.

One problem with the HMICFRS data are that they are produced in a slightly
different table format *almost every year*. The data have been pre-processed to
convert the assortment of CSV, Excel and ODS files on their website to a series
of flat CSV files. These files are not [tidy](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html), 
so some individual wrangling will be required.

Some of the HMICFRS files contain many different variables, and in some cases
the manual pre-processing has to be done for each variable. To minimise this
work (and any consequent mistakes), the only variables that are needed are:

  * police force area population,
  * number of police officers, and
  * workforce by function.

Passive provision should be made to include all functions, just in case they are 
needed for calculating the number of uniformed/front-line officers. Some years 
of data don't include the detailed functions, so only high-level functions (e.g. 
'roads policing') need be included.

HMICFRS uses data from the year prior to the report, so the data listed on their 
website as from 2010 is actually data from 2009.

```{r}
hmic <- list()

# 2010 data
hmic[["2010"]] <- read_csv("../original_data/hmic_data_2011.csv") %>% 
  select(force = "Police force", 
         officers = "Home Office published data 2009-10_FTE equivalent staff_Police Officers",
         roads_officers = "Workforce by function FTE_Police Officers (as at 31/03/2010)_Roads policing") %>% 
  mutate(force = str_replace_all(tolower(force), " & ", " and "), year = 2010)

# 2011 data
hmic[["2011"]] <- read_csv("../original_data/hmic_data_2012.csv") %>% 
  rename(officer_function = Function) %>% 
  rename_all(tolower) %>% 
  mutate(officer_function = tolower(officer_function)) %>% 
  filter(officer_function %in% c("total road policing", "level 2 summary")) %>% 
  gather("force", "value", -officer_function) %>% 
  spread(officer_function, value) %>% 
  rename(officers = `level 2 summary`, 
         roads_officers = `total road policing`) %>% 
  mutate(force = str_replace_all(tolower(force), " & ", " and "), year = 2011)

# 2012—2016 data
# note that the 2016 data have an asterisk after some category names, which have
# been removed during pre-processing so that the file can be processed along
# with the others
hmic_list <- lapply(2012:2016, function (x) {
  read_csv(paste0("../original_data/hmic_data_", (x + 1), ".csv")) %>% 
    rename_all(tolower) %>% 
    rename(poa_category = `poa category`) %>% 
    mutate(poa_category = tolower(poa_category)) %>% 
    filter(poa_category %in% c("total road policing", "level 2 summary")) %>%
    gather("key", "value", -poa_category) %>%
    spread(poa_category, value) %>%
    rename(force = key, officers = `level 2 summary`,
           roads_officers = `total road policing`) %>%
    mutate(force = str_replace_all(force, " & ", " and "), year = x)
})

# merge the individual years with the combined years
bind_rows(c(hmic, hmic_list)) %>% 
  mutate(force = if_else(force == "metropolitan police", "metropolitan", 
                         force)) %>% 
  write_csv("../analysis_data/data_officer_counts.csv", na = "")
rm(hmic, hmic_list)

```


# Historical police workforce data

These data are needed for a chart of the number of police per head since the 
Second World War.


# Historical population data


# Collision data


# Vehicle kilometres data

Estimates of millions of vehicle-kilometres travelled are published for top-tier
local authorities for each year.

These estimates can be converted to police-force areas (PFAs), except that the 
ONS lookup table for PFAs only links to local-authority districts. This is fine
for unitary authorities, but not for two-tier areas. To get around this we can
create a lookup using the ONS lookup file for counties and districts.

```{r}
# load traffic data
traffic_data_la <- read_csv("../original_data/traffic_data.csv") %>% 
  gather("year", "million_km", -`LA Code`, -`Local Authority`) %>% 
  mutate(year = as.integer(year)) %>% 
  # the next line corrects two problems in the traffic data file
  mutate(
    `LA Code` = if_else(`LA Code` == "E06000048", "E06000057", `LA Code`),
    `LA Code` = if_else(`LA Code` == "E08000020", "E08000037", `LA Code`)
  ) %>%
  filter(year >= 2010)

# load LAD->PFA lookup table
pfa_lad_lookup <- read_csv("../analysis_data/lookup_pfa_to_lad.csv") %>% 
  select(LAD16CD, force = PFA16NM) %>% 
  mutate(force = str_replace_all(force, " & ", " and "))

# create CTY->PFA lookup table
pfa_cty_lookup <- read_csv("../analysis_data/lookup_cty_to_lad.csv") %>% 
  left_join(pfa_lad_lookup, by = c("LAD16CD" = "LAD16CD")) %>% 
  group_by(CTY16CD) %>%
  summarise(force = first(force))

# join PFA name to traffic data
traffic_data_la <- traffic_data_la %>% 
  left_join(pfa_lad_lookup, by = c("LA Code" = "LAD16CD")) %>% 
  left_join(pfa_cty_lookup, by = c("LA Code" = "CTY16CD")) %>% 
  mutate(force = if_else(is.na(force.x), force.y, force.x)) %>% 
  select(-force.x, -force.y) %>% 
  filter(!is.na(force))

# sum traffic counts for PFAs
traffic_data_pfa <- traffic_data_la %>% 
  group_by(year, force) %>% 
  summarise(million_km = sum(million_km))

# write CSV file
traffic_data_pfa %>% 
  mutate(force = case_when(
    force == "London, City of" ~ "City of London",
    force == "Metropolitan Police" ~ "Metropolitan",
    TRUE ~ force
  )) %>% 
  write_csv("../analysis_data/data_traffic.csv", na = "")

rm(pfa_cty_lookup, pfa_lad_lookup, traffic_data_la, traffic_data_pfa)
```


# Combining the data

Each of the sections above outputs a tidy CSV file. We can now combine these to
produce the final data file that will be used for analysis.

