---
title: "Data analysis"
output:
  html_notebook: default
---

# Load data and create new variables for analysis

All of the data imported into this file are counts (of officers, FPNs etc.). For
the analysis, some variables need to be converted to different formats:

  * traffic officers should be expressed as a proportion of all officers, and
  * FPNs and breath tests should be expressed per officer.
  
This will allow these variables to be included in models along with the number
of officers in each force.

Since the Met is so much bigger than other forces, it's necessary to convert the
counts of accidents into a rate per 10 million vehicle kilometres travelled. 
This avoids skewing the residuals of the regression models by having such a 
large outlier in the data.

Finally, we need to convert `force` from a character to a factor.

```{r}
d <- read_csv("../analysis_data/traffic_police_and_collisions.csv") %>% 
  mutate(
    roads_officers_prop = roads_officers / officers,
    fpn_all_rate = fpn_all / officers,
    fpn_phone_rate = fpn_phone / officers,
    fpn_speed_rate = fpn_speed / officers,
    breath_test_rate = breath_tests  /officers,
    acc_all_rate = acc_all / (million_km / 10),
    acc_ksi_rate = acc_ksi / (million_km / 10),
    acc_drink_rate = acc_drink / (million_km / 10),
    acc_phone_rate = acc_phone / (million_km / 10),
    acc_speed_rate = acc_speeding / (million_km / 10),
    police_force = as.factor(force)
  )
```

## Centering variables

To ease interpretation of the intercept terms in the models below, we should
mean-center some of the variables. Otherwise the intercept term will be the 
rate of accidents expected if there were no police etc.

```{r}
d <- d %>% mutate(
  officers_mctr = scale(officers, center = TRUE, scale = FALSE),
  roads_officers_prop_mctr = scale(roads_officers_prop, center = TRUE, 
                                   scale = FALSE),
  fpn_all_rate_mctr = scale(fpn_all_rate, center = TRUE, scale = FALSE),
  fpn_phone_rate_mctr = scale(fpn_phone_rate, center = TRUE, scale = FALSE),
  fpn_speed_rate_mctr = scale(fpn_speed_rate, center = TRUE, scale = FALSE),
  breath_test_rate_mctr = scale(breath_test_rate, center = TRUE, scale = FALSE)
)
```


## Accounting for types of force

Analysis of the residuals of models of the form `accidents ~ officers + …`
showed patterns in the residuals by force, which appeared to indicate how urban
the forces were. To account for this, a variable is included indicating the 
type of force (urban or rural) using the classifications shown in footnotes on
page 176 of [Higgins et al (2010)](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/116360/hosb1210-chap7.pdf).

```{r}
d <- d %>% 
  mutate(force_type = case_when(
    force %in% c("Metropolitan") ~ "Metropolitan",
    force %in% c("Greater Manchester", "West Midlands") ~ "large",
    TRUE ~ "other"
  )) %>% arrange(year, force)
  # mutate(force_type = case_when(
  #   force %in% c("Cleveland", "Greater Manchester", "Hertfordshire", 
  #                "Lancashire", "City of London", "Merseyside", "Metropolitan",
  #                "Northumbria", "Nottinghamshire", "South Yorkshire", "Surrey",
  #                "West Midlands", "West Yorkshire") ~ "urban",
  #   TRUE ~ "rural"
  # )) %>% arrange(year, force)
```



Exploratory analysis shows that:

  * City of London has very high accident rates relative to vehicle kilometres,
  * The Met comes out as a substantial outlier in any models, probably because
    of its size and potentially also because it is likely that the relationships
    between the variables in London is different from those in the rest of the
    country.

To prevent these forces skewing the models, they are excluded here.

```{r}
d <- d %>% filter(!force %in% c("City of London", "Metropolitan"))
```


# Checking model assumptions

GLS regression requires a normally distributed response variable, so first we
should check this. We're primarily interested in KSIs, so this is the data we'll
check.

```{r}
hist(d$acc_ksi_rate, breaks = 100)
```


# Model preparation

This list will store the results of all the models run below.

```{r}
m <- list()
```

We need an empty model to compare our models to. Note that `all_` refers to
models predicting *all* collisions, regardless of severity or contributory
factors, while `ksi_` refers to collisions in which somewas was Killed or
Seriously Injured.

In these models, `corAR1()` specifies that the data have first-order
autocorrelation, with `1 | force` saying that the data are grouped by force.
`varIdent(form = ~ 1 | year)` specifies that the variance in the data may be 
different across years. `anova()` tests showed that these specifications fit
the data better than models without these terms.

```{r}
m[["ksi_empty"]] <- gls(
  acc_ksi_rate ~ 1, 
  data = d, 
  correlation = corAR1(form = ~ 1 | police_force),
  weights = varIdent(form = ~ 1 | year),
  method = "ML"
)
summary(m$ksi_empty)
```


# Collisions vs officers

This first model predicts collisions using number of officers.

```{r}
m[["ksi_ofc"]] <- update(m$ksi_empty, . ~ force_type + officers_mctr)
summary(m$ksi_ofc)
```

From this we can see that the co-efficient for `officers_mctr` is significant
but very small: each additional officer decreases the rate of accidents by
-2.05^-5 per million vehicle kilometres.


# Collisions vs officers and traffic officers

We can add the proportion of officers who are traffic officers to the model to 
see if officers with training and experience in traffic matters has any effect.

```{r}
m[["ksi_ofc_tpc"]] <- update(m$ksi_ofc, . ~ . + roads_officers_prop_mctr)
summary(m$ksi_ofc_tpc)
```

The co-efficient for `roads_officers_prop_mctr` is larger than for 
`officers_mctr` but non-significant. 


# Collisions vs officers, traffic officers and activity

We're interested in two types of traffic policing activity: FPNs issued (by 
officers, not by fixed automatic cameras) and breath tests.

```{r}
m[["ksi_ofc_tpc_act"]] <- update(m$ksi_ofc_tpc, 
  . ~ . + fpn_all_rate_mctr + breath_test_rate_mctr)
summary(m$ksi_ofc_tpc_act)
```

```{r}
tibble(
  residuals = residuals(m$ksi_ofc, type = "pearson"), 
  values = fitted.values(m$ksi_ofc), 
  force_type = d$force_type, 
  year = d$year
) %>% 
  ggplot(aes(x = values, y = residuals, group = force_type, colour = force_type)) + 
    geom_point() 
```

